Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-04 paper.qmd
method - impact probability - some editorial tweaks

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.

@@

## Impact probability for a single flight and single debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

where $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
$\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$ and $\varepsilon_k$ is the effective exposed area,
i.e. the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

Using H3 to discretise the area of interest $a$ using H3 cells gives

@@

Where for each cell $h$ the following applies

>>>>

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities resulting in higher numbers of re-entries at the end of the mission lifespan or orbital decay. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.

@@

## Impact probability for a single flight and single debris

Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a transversing flight.
This probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

where 
* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$, and 
* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

Our approach builds on the discretisation of the geo-spatial risk area.
For this purpose we use a hexagon-based tesselation pattern.
Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
Using H3 to discretise the area of interest $a$ gives

@@

The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following
